---
title: "lcwgs-pca-exploration"
author: "diana baetscher"
date: "2024-01-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading in .cov files that Laura Timm generated and playing with different PCAs for my AMSS presentation.


```{r}
######################################
### INSTALL PACKAGES & LOAD FUNCTIONS
packages_needed <- c("ggplot2", "plotly", "ggpubr", "tidyverse")

for(i in 1:length(packages_needed)){
  if(!(packages_needed[i] %in% installed.packages())){install.packages(packages_needed[i])}
  library(packages_needed[i], character.only = TRUE)
}

g_legend <- function(myggplot){
  tmp <- ggplot_gtable(ggplot_build(myggplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}
######################################

```


```{r}
######################################
## DIRECTORIES AND FILE NAMES
DATADIR <- "../lcwgs_data/"
BAM_LIST <- "SALU-7164_filtered_bamslist.txt"
#SAMPLEMETADATA <- "BFALsamplePops.csv"
#METADATA <- "fst_meta_data.txt"
######################################
```




```{r}
### Genome-wide PCA

# read in the covariance matrix from pcangsd
pca <- as.matrix(read.table("../lcwgs_data/SALU-7164-batch-q25_wholegenome-polymorphic.cov"))

# compute eigenvalues from the covariance matrix
pca_e <- eigen(pca)

# extract the first three PCs
first3pcs <- data.frame(pca_e$vectors[,1:3])
dim(pca)

# still need the identities of the 337 samples in the covariance matrix
pca_df <- as.data.frame(first3pcs) %>%
  rename(PC1 = X1, PC2 = X2, PC3 = X3)

```

```{r quick-plot}
pca_df %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_bw()

```

That looks identical to Laura's. Let's see if we can add in some metadata - first for the samples from Jacek's study.

```{r metadata-and-bamlist}
bamlist <- read_tsv("../lcwgs_data/SALU-7164-batch-q25_filtered_bamslist.txt", col_names = F) %>%
  separate(X1, c(NA, NA, NA, NA, NA, "file"), sep = "/") %>%
  separate(file, into = c(NA, "ABLG", NA, NA, NA, NA), sep = "_")

bamlist

# metadata
locs <- read_csv("../lcwgs_data/pop_locs.csv") %>%
  select(-`...6`)

grp_info <- locs %>%
  mutate(original_pop = ifelse(Population == "larval-SEAK", pop_assign, NA)) %>%
  mutate(original_pop = ifelse(str_detect(original_pop, "B"), "B", original_pop)) # Jacek only had four groups

# add info to bamlist samples ***BAMLIST IS THE CANONICAL ORDER FOR THE PCA COVARIANCE MATRIX
sample_ids <- bamlist %>%
  left_join(., grp_info)
  
```
```{r add-group-info-to-pcas}
pca_w_meta <- sample_ids %>%
  bind_cols(., pca_df)

no_jacek <- pca_w_meta %>%
  filter(is.na(original_pop))

jacek <- pca_w_meta %>%
  filter(!is.na(original_pop))

# plot with Jacek's samples superimposed on lcWGS data
  ggplot() +
  geom_point(data = no_jacek, aes(x = PC1, y = PC2, color = original_pop), size = 1.5, alpha = 0.3) +
  geom_point(data = jacek, aes(x = PC1, y = PC2, color = original_pop), size = 3.5, alpha = 0.7) +
  theme_bw() +
  scale_color_manual(values = c("darkcyan", 
                                "deepskyblue", 
                                "salmon", 
                                "darkblue"),
                                na.value = "gray70") +
    labs(
      color = "RAD-seq \ngenetic group",
      x = "PC1 - 20.0%",
      y = "PC2 - 7.9%"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14)

      
    )
    
ggsave("pdf_outputs/Jacek_groups_PCA_lcwgs_data.png", width = 6, height = 4)
```

```{r add-group-info-to-pcas}
pca_w_meta <- sample_ids %>%
  bind_cols(., pca_df)

no_jacek <- pca_w_meta %>%
  filter(is.na(original_pop))

jacek <- pca_w_meta %>%
  filter(!is.na(original_pop))

# plot without Jacek's samples
no_jacek %>%
  filter(pop_assign != "B") %>%
  ggplot() +
  geom_point(aes(x = PC1, y = PC2, color = pop_assign), size = 2, alpha = 0.5) +
 # geom_point(data = jacek, aes(x = PC1, y = PC2, color = original_pop), size = 3.5, alpha = 0.7) +
  theme_bw() +
  scale_color_manual(values = c("darkcyan",
                                "maroon",
                                "goldenrod1",
                                "deepskyblue",
                                "dodgerblue3",
                                "salmon",
                                "darkblue"),
                                na.value = "gray70") +
    labs(
      color = "lcWGS \ngenetic group",
      x = "PC1 - 20.0%",
      y = "PC2 - 7.9%"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14)
    ) + guides(color = guide_legend(override.aes = list(alpha = 1)))

    
ggsave("pdf_outputs/lcwgs_PCA_groups.png", width = 6, height = 4)
```


Now add color to the rest of the PCA samples

```{r all-samples-color-coordinated}
pca_w_meta %>%
  mutate(grp = ifelse(PC1 < -0.04 & PC2 < -0.02 & PC2 > -0.05, NA, pop_assign)) %>%
  #filter(!is.na(pop_assign)) %>%
  mutate(grp = ifelse(str_detect(pop_assign, "B"), "B", grp)) %>%
  ggplot() +
  geom_point(aes(x = PC1, y = PC2, color = grp), size = 2, alpha = 0.5) +
  theme_bw() +
  scale_color_manual(values = c("darkcyan", 
                                "deepskyblue", 
                                "salmon", 
                                "darkblue"),
                     na.value = "gray70") +
    labs(
      color = "lcWGS \ngenetic group",
      x = "PC1 - 20.0%",
      y = "PC2 - 7.9%"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14)
    ) + guides(color = guide_legend(override.aes = list(alpha = 1)))

ggsave("pdf_outputs/lcwgs_data_wes_groups.png", width = 6, height = 4)
```
It's unclear how some of those were assigned groups based on this data. Instead, let's color the unambiguous samples and leave some gray.

What about PC 3??

```{r}
pca_w_meta %>%
  filter(!is.na(pop_assign)) %>%
  mutate(pop_assign = ifelse(str_detect(pop_assign, "B"), "B", pop_assign)) %>%
  filter(pop_assign != "B") %>%
  ggplot() +
  geom_point(aes(x = PC2, y = PC3, color = pop_assign), size = 2, alpha = 0.5) +
  theme_bw() +
  scale_color_manual(values = c("darkcyan", 
                                #"deepskyblue", 
                                "salmon", 
                                "darkblue")) +
    labs(
      color = "lcWGS \ngenetic group",
      x = "PC2 - 7.9%",
      y = "PC3 - %"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14)
    ) + guides(color = guide_legend(override.aes = list(alpha = 1)))



```




```{r plot-w-all-lcwgs-sample-groups}
# What about color based on location information?
# since our hypothesis was structure based on location

df_w_locs <- pca_w_meta %>%
  filter(!is.na(pop_assign) &
           is.na(original_pop)) %>%
  mutate(Population = ifelse(Population == "Southeast Bering Sea", "SE Bering Sea", Population)) %>%
    mutate(Population = ifelse(Population == "cGOA", "Central GOA", Population)) %>%
      mutate(Population = ifelse(Population == "eGOA", "Eastern GOA", Population)) %>%
      mutate(Population = ifelse(Population == "wGOA", "Western GOA", Population))
  
df_w_locs$Population <- factor(df_w_locs$Population, levels = c("Aleutian Islands", "Central Bering Sea", "SE Bering Sea", "Western GOA", "Central GOA", "Eastern GOA", "SEAK", "South"))
  
loc_plot1 <- df_w_locs %>%  
  mutate(pop_assign = ifelse(str_detect(pop_assign, "B"), "B", pop_assign)) %>%
  ggplot() +
  geom_point(aes(x = PC1, y = PC2, color = Population), size = 1, alpha = 0.6) +
  theme_bw() +
  # scale_color_manual(values = c("darkcyan", 
  #                               #"deepskyblue", 
  #                               "salmon", 
  #                               "darkblue")) +
    labs(
      color = "Sampling location",
      x = "PC1 - 20.0%",
      y = "PC2 - 7.9%%"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.position = "none"
      #legend.title = element_text(size = 14),
      #legend.text = element_text(size = 12)
    ) #+ guides(color = guide_legend(override.aes = list(alpha = 1)))


loc_plot1

ggsave("pdf_outputs/lcwgs_sampling_locations.png", width = 6, height = 4)

```
What's the geographic difference between Eastern GOA and SE AK?

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

goa_map <- ggplot(data = world) +
  geom_sf() +
  coord_sf(xlim = c(-179, -120), 
           ylim = c(62, 35), 
           expand = FALSE) +
  theme_bw()

loc_plot2 <- goa_map +
  geom_point(data = df_w_locs, aes(x = StartLongitudeDD, y = StartLatitudeDD, color = Population)) +
  labs(
      color = "Sampling location",
      x = "Longitude",
      y = "Latitude"
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)
    ) +
  scale_x_continuous(breaks = c(-170, -150, -130)) +
  scale_y_continuous(breaks = c(35, 45, 55))
  
loc_plot2


goa_map +
  geom_point(data = df_w_locs, aes(x = StartLongitudeDD, y = StartLatitudeDD, color = Population)) +
  labs(
      color = "Sampling location",
      x = "",
      y = ""
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)
    ) +
  scale_x_continuous(breaks = c(-170, -150, -130)) +
  scale_y_continuous(breaks = c(35, 45, 55))

ggsave("pdf_outputs/lcwgs_sampling_locations.png", width = 6, height = 4)
```

```{r}
library(patchwork)

loc_plot1 + loc_plot2 + plot_layout(guides = "collect")

ggsave("pdf_outputs/lcwgs_sampling_locations_pca.png", width = 10, height = 8)
```



Plot genetic groups (rather than sampling locations) on map

```{r genetic-assignments-lcWGS}
tmp <- df_w_locs %>%
  filter(pop_assign != "B")
ā
tmp$pop_assign <- factor(tmp$pop_assign, levels = c("A", "B1", "B2", "C", "B3", "B4", "D"))

goa_map +
  geom_point(data = tmp, aes(x = StartLongitudeDD, y = StartLatitudeDD, color = pop_assign), size = 2.5, alpha = 0.8) +
  facet_wrap(~pop_assign) +
  theme_minimal() +
  labs(
      color = "Genetic group",
      x = "",
      y = ""
    ) +
    theme(
      axis.title.x = element_text(margin = margin(t = 10), size = 14),
            axis.title.y = element_text(margin = margin(r = 10), size = 14),
      # legend.title = element_text(size = 14),
      # legend.text = element_text(size = 12),
      strip.text = element_text(size = 14),
      legend.position = "none"
    ) +
  scale_x_continuous(breaks = c(-170, -150, -130)) +
  scale_y_continuous(breaks = c(35, 45, 55)) +
  scale_color_manual(values = c("darkcyan",
                                "maroon",
                                "goldenrod1",
                                "salmon",
                                "deepskyblue",
                                "dodgerblue3",
                                "darkblue"),
                                na.value = "gray70")

ggsave("pdf_outputs/lcwgs_pop_assign_map_grid.png", width = 7, height = 6)
```

